---
title: "Analysing a dose-response experiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing a dose-response experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Code in this vignette is illustrative and is not executed: it points at plate
# files that live outside the package. Copy the patterns below and substitute
# your own paths.
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This vignette walks through a complete dose-response analysis with `drcanalyser`,
using an ARID1A-knockout viability experiment as a worked example. The goal is to
compare drug sensitivity between a control (`Ctrl`) and knockout (`KO`) cell line,
each measured across three biological replicates.

The pipeline has four steps:

1. **Describe** each plate as a list of file paths and options.
2. **Process** the replicate plates of a condition into a single fitted
   dose-response model with `process_plates()`.
3. **Plot** one or more conditions together with `plot_drc()`.
4. **Save** the figure and statistics with `save_results()`.

```{r}
library(drcanalyser)
```

## 1. Describing a plate

Every plate is a named list of arguments passed through to `process_plate()`.
At minimum you supply `main_path` — the SparkControl Excel export for the
treatment plate. For a GI50 / growth-rate analysis you also give `gi50_path`,
the matching day-zero plate.

```{r}
ctrl_rep1 <- list(
  main_path = "data/EFO_Ctrl_rep1_main.xlsx",
  gi50_path = "data/EFO_Ctrl_rep1_GI50.xlsx"
)
```

### Excluding wells

Mis-dosed or unreliable wells are dropped with `exclude`, a list of
`c(row, column)` coordinates. Here we drop the top-dose column because it was
pipetted incorrectly:

```{r}
ctrl_rep2 <- list(
  main_path = "data/EFO_Ctrl_rep2_main.xlsx",
  gi50_path = "data/EFO_Ctrl_rep2_GI50.xlsx",
  exclude = list(c("B", 2), c("C", 2), c("D", 2))
)
```

### Plate layouts

`drcanalyser` ships default layouts that map each well to a role (`trt`,
`neg_ctrl`, `bckgrnd`, `gi50`) and a concentration — one for IC50 plates and one
for GI50 plates. These defaults are used automatically, so most plates need no
layout argument.

If a plate used a non-standard layout, point `plate_layout` at your own layout
workbook (an Excel file with `layout`, `concs` and optional `gi50` sheets — see
`?read_plate_layout`):

```{r}
ctrl_rep3 <- list(
  main_path = "data/EFO_Ctrl_rep3_main.xlsx",
  gi50_path = "data/EFO_Ctrl_rep3_GI50.xlsx",
  plate_layout = "data/plate_layout_custom.xlsx",
  exclude = list(c("B", 2), c("C", 2), c("D", 2))
)
```

## 2. Processing replicates into a model

`process_plates()` takes a list of replicate plate specs, normalises and
combines them, and fits a single three-parameter log-logistic dose-response
curve (`drc::LL.3u`). Choose how intensities are normalised with
`normalisation_method`:

- `"IC"` — relative viability (`intensity / neg_ctrl`); for IC50.
- `"GI"` — growth inhibition against the day-zero baseline; for GI50.
- `"GR"` — growth-rate normalisation (Hafner *et al.*, 2016).

`assay_id` labels the condition and becomes the grouping/legend key downstream.

```{r}
efo_ctrl <- process_plates(
  list(ctrl_rep1, ctrl_rep2, ctrl_rep3),
  assay_id = "EFO21-Ctrl",
  normalisation_method = "GR"
)
```

The knockout condition is described and processed the same way:

```{r}
efo_ko <- process_plates(
  list(
    list(
      main_path = "data/EFO_KO_rep1_main.xlsx",
      gi50_path = "data/EFO_KO_rep1_GI50.xlsx",
      exclude = list(c("B", 2), c("C", 2), c("D", 2))
    ),
    list(
      main_path = "data/EFO_KO_rep2_main.xlsx",
      gi50_path = "data/EFO_KO_rep2_GI50.xlsx",
      exclude = list(c("B", 2), c("C", 2), c("D", 2))
    ),
    list(
      main_path = "data/EFO_KO_rep3_main.xlsx",
      gi50_path = "data/EFO_KO_rep3_GI50.xlsx",
      exclude = list(c("B", 2), c("C", 2), c("D", 2))
    )
  ),
  assay_id = "EFO21-KO",
  normalisation_method = "GR"
)
```

Each result is a list with the combined `$plate` data frame and the fitted
`$model`.

## 3. Plotting and comparing conditions

`plot_drc()` takes a *list* of processed conditions and overlays their curves on
a single log-concentration axis. It returns the ggplot in `$plot` and a table of
ED10/50/90 statistics in `$stats`.

```{r}
result <- plot_drc(list(efo_ko, efo_ctrl))
result$plot
result$stats
```

Set `plot_mean = TRUE` to collapse replicates at each concentration to a mean
with standard-error bars:

```{r}
result_mean <- plot_drc(list(efo_ko, efo_ctrl), plot_mean = TRUE)
```

## 4. Saving results

`save_results()` writes the plot as an pdf and the statistics as a CSV into a
folder, with an optional filename suffix:

```{r}
save_results(
  plot_drc(list(efo_ko, efo_ctrl)),
  save_folder = "figures",
  append_file_name = "EFO21_combined"
)
```

## Derived metrics

Because each processed plate carries its raw control means, you can derive
further quantities directly from the `$plate` data frames. For example, the
average growth ratio (untreated control over day-zero, `XCTRL / X0`) per plate:

```{r}
library(dplyr)

agr <- rbind(efo_ctrl$plate, efo_ko$plate) %>%
  select(ASSAY_ID, PLATE_ID, X0, XCTRL) %>%
  distinct() %>%
  mutate(AGR = XCTRL / X0)
```

## Summary

A full experiment is just these steps repeated per cell line or drug: describe
each plate as a list, `process_plates()` the replicates of a condition, then
`plot_drc()` the conditions you want to compare. See the individual function
help pages (e.g. `?process_plates`, `?plot_drc`) for the complete set of
arguments.
